
## Objective

This report performs a thorough non-graphical exploratory data analysis (EDA) of the BodyFat dataset.

Key requirement addressed immediately:
- All columns are numeric.
- `Density` is a direct leakage variable because `BodyFat` is mathematically derived from it using Siri's equation:

$$
\text{BodyFat} = \frac{495}{\text{Density}} - 450
$$

Because of this exact deterministic relationship, `Density` must be removed before any realistic predictive modeling.

## Setup and Data Loading

```{r}
library(tidyverse)

bodyfat_raw <- read_csv("data/bodyfat.csv", show_col_types = FALSE)
```

## Data Structure Inspection (glimpse and summary)

```{r}
# Compact structure view
bodyfat_raw %>% glimpse()

# Numeric summaries for each feature
summary(bodyfat_raw)
```

### Immediate Leakage Check for Density

```{r}
# Verify Siri equation numerically against observed BodyFat values
bodyfat_with_check <- bodyfat_raw %>%
  mutate(
    BodyFat_from_Density = 495 / Density - 450,
    Leakage_abs_error = abs(BodyFat - BodyFat_from_Density)
  )

bodyfat_with_check %>%
  summarise(
    max_abs_error = max(Leakage_abs_error),
    mean_abs_error = mean(Leakage_abs_error),
    min_abs_error = min(Leakage_abs_error)
  )
```

Interpretation:
- Near-zero reconstruction error confirms `Density` contains direct information about the target `BodyFat`.
- Any model using `Density` would be unrealistically optimistic (target leakage).
- For all downstream analysis intended to support modeling, use a dataset with `Density` removed.

```{r}
bodyfat <- bodyfat_raw %>% select(-Density)
```

## Dataset Dimensions, Types, and Duplicates

```{r}
# Shape
c(rows = nrow(bodyfat_raw), columns = ncol(bodyfat_raw))

# Confirm all columns are numeric in raw data
sapply(bodyfat_raw, class)

# Duplicate rows
bodyfat_raw %>%
  summarise(duplicate_rows = sum(duplicated(.)))
```

## Descriptive Statistics (Post-Leakage Removal)

```{r}
# Summary table for all original features (including Density for EDA completeness)
summary_table_all_features <- map_dfr(names(bodyfat_raw), function(v) {
  x <- bodyfat_raw[[v]]
  mean_x <- mean(x, na.rm = TRUE)
  sd_x <- sd(x, na.rm = TRUE)

  tibble(
    feature = v,
    mean = mean_x,
    median = median(x, na.rm = TRUE),
    sd = sd_x,
    iqr = IQR(x, na.rm = TRUE),
    cv = ifelse(mean_x == 0, NA_real_, sd_x / mean_x)
  )
})

summary_table_all_features
```

## Missing Value Analysis

```{r}
# Missingness by feature (count and percent)
missing_by_feature <- bodyfat_raw %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "missing_n"
  ) %>%
  mutate(missing_pct = 100 * missing_n / nrow(bodyfat_raw)) %>%
  arrange(desc(missing_n), feature)

missing_by_feature
```

```{r}
# Missingness by observation (row-wise NA count)
row_missing <- bodyfat_raw %>%
  mutate(row_id = row_number(), missing_n = rowSums(is.na(across(everything())))) %>%
  select(row_id, missing_n)

row_missing %>%
  summarise(
    complete_rows = sum(missing_n == 0),
    incomplete_rows = sum(missing_n > 0),
    pct_complete_rows = 100 * mean(missing_n == 0),
    max_missing_in_a_row = max(missing_n)
  )
```

```{r}
# Distribution of row-level missingness severity
row_missing %>%
  count(missing_n, name = "row_count") %>%
  arrange(missing_n)
```

## Implausible Values and Data Quality Flags

```{r}
# Flag records that are statistically and/or physically implausible
implausible_rows <- bodyfat_raw %>%
  mutate(row_id = row_number()) %>%
  filter(BodyFat <= 1 | Height < 55)

implausible_rows %>%
  select(row_id, BodyFat, Height, Age, Weight, Density, everything())
```

```{r}
# Compact report of the two specific issues requested
implausible_summary <- bodyfat_raw %>%
  mutate(row_id = row_number()) %>%
  summarise(
    bodyfat_zero_row = row_id[which.min(BodyFat)],
    bodyfat_zero_value = min(BodyFat, na.rm = TRUE),
    low_height_row = row_id[which.min(Height)],
    low_height_value = min(Height, na.rm = TRUE)
  )

implausible_summary
```

Interpretation and treatment recommendation:
- The row with `BodyFat` approximately 0% (exactly 0.0 in this file) is biologically very unlikely for this population and should be treated as a suspect observation.
- The row with `Height = 29.5` inches is not plausible for an adult male body-composition dataset and is most likely a data entry/coding error (for example, a dropped leading digit such as 69.5).
- Decision guideline:
  - If source records can be audited, correct clear typos (preferred).
  - If audit is not possible, mark these values as invalid (`NA`) and use robust handling/imputation, or remove the affected row(s) in sensitivity analysis.
  - Do not keep these records unchanged in a final modeling dataset; they are better treated as probable data errors than natural outliers.

## Outlier Diagnostics by IQR Rule (Non-Graphical)

```{r}
outlier_counts <- map_dfr(names(bodyfat), function(v) {
  x <- bodyfat[[v]]
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr

  tibble(
    variable = v,
    lower_bound = lower,
    upper_bound = upper,
    outlier_n = sum(x < lower | x > upper, na.rm = TRUE),
    outlier_pct = 100 * mean(x < lower | x > upper, na.rm = TRUE)
  )
})

outlier_counts %>% arrange(desc(outlier_n))
```

## Correlation Analysis (No Plots)

```{r}
# Correlation matrix after dropping Density
cor_matrix <- cor(bodyfat, use = "pairwise.complete.obs")
cor_matrix
```

```{r}
# Rank strongest absolute correlations with BodyFat
target_cor <- tibble(
  variable = setdiff(colnames(bodyfat), "BodyFat"),
  cor_with_bodyfat = cor_matrix[setdiff(colnames(bodyfat), "BodyFat"), "BodyFat"]
) %>%
  mutate(abs_cor = abs(cor_with_bodyfat)) %>%
  arrange(desc(abs_cor))

target_cor
```

```{r}
# Top overall pairwise correlations (excluding self-pairs and duplicates)
cor_long <- as.data.frame(as.table(cor_matrix)) %>%
  as_tibble() %>%
  rename(var1 = Var1, var2 = Var2, correlation = Freq) %>%
  filter(var1 < var2) %>%
  mutate(abs_correlation = abs(correlation)) %>%
  arrange(desc(abs_correlation))

head(cor_long, 15)
```

```{r}
# All feature pairs with strong linear association (|correlation| > 0.8)
high_cor_pairs <- cor_long %>%
  filter(abs_correlation > 0.8) %>%
  arrange(desc(abs_correlation))

high_cor_pairs
```

```{r}
# Focus on circumference clusters requested in the assignment
focus_pairs <- tibble(
  var1 = c("Chest", "Chest", "Abdomen", "Knee", "Knee", "Ankle"),
  var2 = c("Abdomen", "Hip", "Hip", "Ankle", "Wrist", "Wrist")
) %>%
  mutate(
    correlation = map2_dbl(var1, var2, ~ cor_matrix[.x, .y]),
    abs_correlation = abs(correlation)
  ) %>%
  mutate(over_0_8 = abs_correlation > 0.8)

focus_pairs
```

Interpretation:
- Use `high_cor_pairs` as the definitive list of all pairs above 0.8.
- In the torso measurements, `Chest`, `Abdomen`, and `Hip` typically form a strong cluster (high positive correlations), indicating overlapping adiposity/body-size information.
- For `Knee`, `Ankle`, and `Wrist`, inspect `focus_pairs`: if `over_0_8` is `FALSE` for most/all combinations, these distal measurements are related but do not form as tight a cluster as torso circumferences.

## Distribution Shape Indicators (Skewness/Kurtosis Without Plots)

```{r}
# Helper functions avoid extra package dependencies
sample_skewness <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 3) return(NA_real_)
  m <- mean(x)
  s <- sd(x)
  if (s == 0) return(0)
  n / ((n - 1) * (n - 2)) * sum(((x - m) / s)^3)
}

sample_excess_kurtosis <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 4) return(NA_real_)
  m <- mean(x)
  s <- sd(x)
  if (s == 0) return(0)
  term1 <- (n * (n + 1)) / ((n - 1) * (n - 2) * (n - 3)) * sum(((x - m) / s)^4)
  term2 <- (3 * (n - 1)^2) / ((n - 2) * (n - 3))
  term1 - term2
}

shape_stats <- map_dfr(names(bodyfat), function(v) {
  x <- bodyfat[[v]]
  tibble(
    variable = v,
    skewness = sample_skewness(x),
    excess_kurtosis = sample_excess_kurtosis(x)
  )
})

shape_stats %>% arrange(desc(abs(skewness)))
```

## Final EDA Conclusions (Non-Graphical)

1. All variables are numeric, as confirmed by structure inspection.
2. `Density` is a direct deterministic transform of `BodyFat` under Siri's equation and is therefore a leakage feature.
3. Any realistic supervised model for `BodyFat` should be built from `bodyfat` (the dataset with `Density` removed).
4. Missing-value analysis indicates no formal `NA` values in this file, so imputation is not required for missingness itself.
5. Data quality checks still identified implausible non-missing values (notably `BodyFat = 0.0` and `Height = 29.5`), which should be treated as probable data errors unless source verification supports them.
6. The `high_cor_pairs` table provides the full list of feature pairs with `|r| > 0.8`, highlighting strong multicollinearity among body-size/circumference measures.
7. The focused circumference check (`focus_pairs`) should be used to report whether torso measures (`Chest`/`Abdomen`/`Hip`) cluster more strongly than distal measures (`Knee`/`Ankle`/`Wrist`) in this dataset.
8. Before modeling, combine these findings into preprocessing decisions: drop leakage (`Density`), handle implausible records, and mitigate multicollinearity (for example by feature selection, regularization, or dimensionality reduction).
