
## Objective

Build and tune two tree-based regression models with tidymodels to predict `BodyFat`:
- Random Forest
- XGBoost

`Density` is dropped at the start because it is a leakage variable.

## Setup

```{r}
library(tidymodels)
library(vip)
library(doParallel)
library(scales)

tidymodels_prefer()
set.seed(2026)

# Parallel backend for faster tuning
cl <- makePSOCKcluster(parallel::detectCores(logical = FALSE))
registerDoParallel(cl)
```

## Data Preparation

```{r}
bodyfat_raw <- readr::read_csv("data/bodyfat.csv", show_col_types = FALSE)

# Drop leakage variable from all further steps
bodyfat <- bodyfat_raw %>%
  select(-Density)

# 75/25 split
set.seed(2026)
bodyfat_split <- initial_split(bodyfat, prop = 0.75)
train_data <- training(bodyfat_split)
test_data <- testing(bodyfat_split)

# Recipe: remove zero variance and normalize all predictors
bodyfat_recipe <- recipe(BodyFat ~ ., data = train_data) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())

bodyfat_recipe
```

## Cross-Validation

```{r}
set.seed(2026)
folds <- vfold_cv(train_data, v = 5)
folds
```

## Random Forest Tuning

```{r}
rf_spec <- rand_forest(
  mode = "regression",
  trees = 1000,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity")

rf_workflow <- workflow() %>%
  add_recipe(bodyfat_recipe) %>%
  add_model(rf_spec)

p <- ncol(train_data) - 1
rf_params <- parameters(
  mtry(range = c(1L, p)),
  min_n(range = c(2L, 30L))
)

rf_grid <- grid_regular(
  rf_params,
  levels = c(mtry = 3, min_n = 3)
)

rf_grid
```

```{r}
set.seed(2026)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

collect_metrics(rf_tune)
```

```{r}
autoplot(rf_tune)
```

## XGBoost Tuning

```{r}
xgb_spec <- boost_tree(
  mode = "regression",
  trees = 1000,
  learn_rate = tune(),
  tree_depth = tune(),
  mtry = tune(),
  min_n = 5,
  loss_reduction = 0,
  sample_size = 1
) %>%
  set_engine("xgboost")

xgb_workflow <- workflow() %>%
  add_recipe(bodyfat_recipe) %>%
  add_model(xgb_spec)

xgb_params <- parameters(
  learn_rate(),
  tree_depth(),
  mtry(range = c(1L, p))
)

set.seed(2026)
xgb_grid <- grid_random(
  xgb_params,
  size = 20
)

xgb_grid
```

```{r}
set.seed(2026)
xgb_tune <- tune_grid(
  xgb_workflow,
  resamples = folds,
  grid = xgb_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

collect_metrics(xgb_tune)
```

```{r}
autoplot(xgb_tune)
```

## Final Evaluation on Test Set

```{r}
# Select best hyperparameters by RMSE
best_rf <- select_best(rf_tune, metric = "rmse")
best_xgb <- select_best(xgb_tune, metric = "rmse")

best_rf
best_xgb
```

```{r}
# Finalize workflows with best hyperparameters
final_rf_workflow <- finalize_workflow(rf_workflow, best_rf)
final_xgb_workflow <- finalize_workflow(xgb_workflow, best_xgb)

# Refit on full training set and evaluate on test set
rf_last <- last_fit(
  final_rf_workflow,
  split = bodyfat_split,
  metrics = metric_set(rmse, mae, rsq)
)

xgb_last <- last_fit(
  final_xgb_workflow,
  split = bodyfat_split,
  metrics = metric_set(rmse, mae, rsq)
)

rf_metrics <- collect_metrics(rf_last) %>% mutate(model = "Random Forest")
xgb_metrics <- collect_metrics(xgb_last) %>% mutate(model = "XGBoost")

bind_rows(rf_metrics, xgb_metrics) %>%
  select(model, .metric, .estimate)
```

```{r}
# Predicted vs actual scatter plot for each model
rf_preds <- collect_predictions(rf_last) %>% mutate(model = "Random Forest")
xgb_preds <- collect_predictions(xgb_last) %>% mutate(model = "XGBoost")

preds_all <- bind_rows(rf_preds, xgb_preds)

ggplot(preds_all, aes(x = BodyFat, y = .pred)) +
  geom_point(alpha = 0.75, color = "#2A9D8F") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#D1495B") +
  facet_wrap(~ model) +
  labs(
    title = "Predicted vs Actual BodyFat on Test Set",
    x = "Actual BodyFat",
    y = "Predicted BodyFat"
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:
- Points closer to the dashed 45-degree line indicate better predictions.
- Compare spread around the line across panels to judge which tuned model generalizes better.

## Variable Importance (Top 10 Features)

```{r}
# Fit finalized workflows on training data so we can inspect model internals
rf_fit <- fit(final_rf_workflow, data = train_data)
xgb_fit <- fit(final_xgb_workflow, data = train_data)

rf_engine <- extract_fit_engine(rf_fit)
xgb_engine <- extract_fit_engine(xgb_fit)
```

```{r}
# Random Forest variable importance
vip::vip(rf_engine, num_features = 10) +
  labs(title = "Random Forest Variable Importance (Top 10)") +
  theme_minimal(base_size = 11)
```

```{r}
# XGBoost variable importance
vip::vip(xgb_engine, num_features = 10) +
  labs(title = "XGBoost Variable Importance (Top 10)") +
  theme_minimal(base_size = 11)
```

```{r}
# Tabular importance to inspect ranking of Abdomen explicitly
rf_vi <- vip::vi(rf_engine) %>% mutate(model = "Random Forest")
xgb_vi <- vip::vi(xgb_engine) %>% mutate(model = "XGBoost")

bind_rows(rf_vi, xgb_vi) %>%
  group_by(model) %>%
  arrange(desc(Importance), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()
```

Discussion:
- `Abdomen` appears at or near rank 1 in both models.
- This aligns with prior EDA that central body circumference has the strongest relationship with body fat.

## Cleanup

```{r}
stopCluster(cl)
registerDoSEQ()
```
